---
title: "Global Diabetes and Kidney Disease Burden Intelligence Report"
author: "KIPKOECH PATRICK"
date: "17 February 2026"
format: pdf
---

\begin{center}
\textbf{Registration Number:} SDS6/48451/2025 \\[1cm]

\textbf{Course Title:} Statistical Computing \\[0.5cm]
\textbf{Course Code:} SDS6103 \\[1cm]

\textbf{Program:} MSc Data Science \\[0.5cm]
\textbf{Faculty:} Faculty of Science and Technology \\[0.5cm]
\textbf{Department:} Department of Mathematics \\[1cm]

\textbf{Lecturer:} Dr. Timothy Kamanu \\[1cm]
\end{center}

\newpage
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(tidyverse)
library(forecast)
library(cluster)
library(knitr)

# Load cleaned dataset (same as app.R)
diabetes_data <- read_csv("data/diabetes_clean.csv", show_col_types = FALSE)
```

\newpage

# 1. Introduction

Non-communicable diseases (NCDs) are a leading cause of global mortality and long-term disability. Among them, diabetes and kidney diseases contribute significantly to Disability-Adjusted Life Years (DALYs), reflecting both premature mortality and years lived with disability.

This report provides a data-driven assessment of global burden patterns, geographic differences, country rankings, clustering behavior, and 5-year forecasts using time-series modeling.

---

# 2. Data Source

The dataset was obtained from **Our World in Data** and contains annual DALY estimates for diabetes and kidney diseases across countries.

Source:  
https://ourworldindata.org/grapher/disease-burden-from-ncds

Only the cleaned dataset used in the Shiny dashboard is included in this submission.

---

# 3. Executive Overview (Global Trend)

```{r}
global_trend <- diabetes_data %>%
  group_by(year) %>%
  summarise(global_total = sum(daly, na.rm = TRUE), .groups = "drop")

ggplot(global_trend, aes(year, global_total)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  labs(
    title = "Global Diabetes & Kidney Disease Burden (Total DALYs)",
    x = "Year",
    y = "Total DALYs"
  ) +
  theme_minimal()
```

**Insight:**  
The global burden demonstrates a persistent upward trajectory, reflecting population growth, aging demographics, and rising lifestyle-related risk factors.

\newpage

# 4. Country Benchmarking

## 4.1 Top 10 Countries (Latest Year)

```{r}
latest_year <- max(diabetes_data$year)

top10 <- diabetes_data %>%
  filter(year == latest_year) %>%
  arrange(desc(daly)) %>%
  slice(1:10)

ggplot(top10, aes(reorder(country, daly), daly)) +
  geom_col(fill = "darkred") +
  coord_flip() +
  labs(
    title = paste("Top 10 Countries by DALYs in", latest_year),
    x = "Country",
    y = "DALYs"
  ) +
  theme_minimal()
```

**Observation:**  
High-population countries dominate total DALYs, reflecting both population size and epidemiological transition effects.

\newpage

## 4.2 Selected Country Comparison

```{r}
selected_countries <- c("United States", "India", "China", "Brazil", "Germany")

diabetes_data %>%
  filter(country %in% selected_countries) %>%
  ggplot(aes(year, daly, color = country)) +
  geom_line(linewidth = 1.1) +
  labs(
    title = "Disease Burden in Selected Countries",
    x = "Year",
    y = "DALYs",
    color = "Country"
  ) +
  theme_minimal()
```

**Insight:**  
Rapidly developing nations exhibit strong growth trends, while high-income countries show steadier but sustained burden levels.

---

# 5. Continental Comparison

continent_summary <- diabetes_data %>%
  filter(!is.na(continent)) %>%
  group_by(continent, year) %>%
  summarise(total = sum(daly, na.rm = TRUE), .groups = "drop")

ggplot(continent_summary, aes(year, total, color = continent)) +
  geom_line(linewidth = 1.1) +
  labs(
    title = "Continental Comparison of Disease Burden",
    x = "Year",
    y = "Total DALYs"
  ) +
  theme_minimal()

**Observation:**  
Asia carries the largest burden, followed by Africa and Europe, reflecting demographic and epidemiological differences.

\newpage

# 6. Country Clustering Analysis

```{r}
latest_data <- diabetes_data %>%
  filter(year == latest_year) %>%
  select(country, daly)

scaled_data <- scale(latest_data$daly)

set.seed(123)
kmeans_model <- kmeans(scaled_data, centers = 3)

latest_data$cluster <- as.factor(kmeans_model$cluster)

ggplot(latest_data, aes(x = cluster, y = daly)) +
  geom_boxplot(fill = "skyblue") +
  labs(
    title = "Country Clusters Based on Disease Burden",
    x = "Cluster",
    y = "DALYs"
  ) +
  theme_minimal()
```

**Insight:**  
Countries naturally group into low-, medium-, and high-burden clusters, indicating structural global health inequalities.

\newpage

# 7. Forecast & Projection (5-Year ARIMA Model)

```{r}
ts_data <- ts(global_trend$global_total, 
              start = min(global_trend$year), 
              frequency = 1)

model <- auto.arima(ts_data)
forecast_5 <- forecast(model, h = 5)

plot(forecast_5)
```

**Forecast Interpretation:**  
The ARIMA model predicts continued upward growth over the next five years, suggesting increasing health system pressure if no major intervention occurs.

---

# 8. Conclusion

This intelligence-driven analysis reveals:

- A sustained global increase in diabetes and kidney disease burden.
- Significant variation across countries and continents.
- Structural clustering of countries into burden tiers.
- A projected continuation of upward trends over the next five years.

The findings emphasize the need for:

- Preventive healthcare policies
- Early detection systems
- Targeted interventions in high-burden regions
- Long-term investment in NCD management

---

# References

Our World in Data (2024). Disease burden from non-communicable diseases.  
https://ourworldindata.org/grapher/disease-burden-from-ncds

World Health Organization (2023). Noncommunicable diseases fact sheet.  
https://www.who.int/news-room/fact-sheets/detail/noncommunicable-diseases